---
author:
title: "EDAV Fall 2019 PSet 2"
output: 
  pdf_document: 
    fig_height: 3
    fig_width: 5
---

Read *Graphical Data Analysis with R*, Ch. 4, 5

Grading is based both on your graphs and verbal explanations. Follow all best practices as discussed in class. Data manipulation should not be hard coded. That is, your scripts should be written to work for new data.

```{r setup, include=FALSE}
 # keep this chunk in your .Rmd file
 knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(tidyverse)
library(robotstxt)
library(rvest)
library(plotly)
```
    
### 1. useR2016! survey

[18 points]

Data: `useR2016` dataset in the **forwards** package (available on CRAN)

For parts (a) and (b):

* Do not toss NAs.
* Do some research to find the wording of the questions asked as relevant and include them in the titles of your graphs.
* Include the dataset name, package name, and link to the question wording source in the graph caption.

(a) Create a horizontal bar chart of the responses to Q20.


(b) Create a vertical bar chart of the responses to Q11.


(c) Create a horizontal stacked bar chart showing the proportion of respondents for each level of Q11 who are over 35 vs. 35 or under. Use a descriptive title. 


(d) Create a horizontal stacked bar chart showing the proportional breakdown of Q11 for each level of Q3, faceted on Q2. Use a descriptive title. 


(e) For the next part, we will need to be able to add line breaks (`\n`) to long tick mark labels. Write a function that takes a character string and a desired approximate line length in number of characters and substitutes a line break for the first space after every multiple of the specified line length.


(f) Create a horizontal bar chart that shows the percentage of positive responses for `Q13 - Q13_F`. Use your function from part (e) to add line breaks to the responses. Your graph should have one bar each for `Q13 - Q13_F`.


### 2. Rotten Tomatoes

[18 points]

To get the data for this problem, we'll use the **robotstxt** package to check that it's ok to scrape data from Rotten Tomatoes and then use the **rvest** package to get data from the web site.

(a) Use the `paths_allowed()` function from **robotstxt** to make sure it's ok to scrape https://www.rottentomatoes.com/browse/box-office/. Then use **rvest** functions to find relative links to individual movies listed on this page. Finally, paste the base URL to each to create a character vector of URLs.

Display the first six lines of the vector.

```{r}
paths_allowed(paths = "https://www.rottentomatoes.com/browse/box-office/", warn = TRUE)
```

```{r}
rottentomatoes.box_office_urls <- function(rank_id = 0, country = 'us') {
  read_html(sprintf("https://www.rottentomatoes.com/browse/box-office/?rank_id=%d&country=%s", rank_id, country)) %>%
    html_node("table.center.table") %>%
    html_nodes("td > a") %>%
    html_attr("href") %>%
    map(~ str_c("https://www.rottentomatoes.com", .)) %>%
    as.character
}
links <- rottentomatoes.box_office_urls()
head(links, 6)
```

(b) Write a function to read the content of one page and pull out the title, tomatometer score and audience score of the film. Then iterate over the vector of all movies using `do.call() / rbind() / lapply()` or `dplyr::bind_rows() / purrr::map()` to create a three column data frame (or tibble).

Display the first six lines of your data frame.

(Results will vary depending on when you pull the data.)
 
For help, see this SO post: https://stackoverflow.com/questions/36709184/build-data-frame-from-multiple-rvest-elements

```{r}
rottentomatoes.get_movies <- function(movie_urls) {
  pager <- function(page) {
      doc <- read_html(url(page))
      data.frame(
          title = doc %>% html_node("div.score_panel > h1") %>% html_text() %>% as.character,
          tomatometer = doc %>% html_node("#tomato_meter_link") %>% html_text() %>% str_trim() %>% str_replace("%","") %>% as.double,
          audience_score = doc %>% html_node("div.audience-score > h2") %>% html_text() %>% str_trim() %>% str_replace("%","") %>% as.double,
          url = page,
          stringsAsFactors=FALSE
      )
  }
  return(do.call(rbind, lapply(movie_urls, pager))) 
}
movies <- rottentomatoes.get_movies(rottentomatoes.box_office_urls())
head(movies, 3)
```

Write your data to file so you don't need to scrape the site each time you need to access it.
```{r}
write.csv(movies, file = "rottentomatoes_movies.csv", row.names=FALSE)
```

(c) Create a Cleveland dot plot of tomatometer scores.
```{r fig.height=4, fig.width=3}
# create a theme for dot plots, which can be reused
theme_dotplot <- theme_bw(12) +
    theme(
        plot.title = element_text(size = rel(1)),
        axis.text.y = element_text(size = rel(.75)),
        axis.ticks.y = element_blank(),
        axis.title.x = element_text(size = rel(.75)),
        panel.grid.major.x = element_line(size = 0.2),
        panel.grid.major.y = element_line(size = 0.75),
        panel.grid.minor.x = element_blank()
    )
        
# create the plot
movies %>%
  # Replace NA score with -1, which causes those movies to be sorted at the bottom of the plot
  # The value of -1 is not shown since we wet the scale for x to be [0,100]
  mutate(tomatometer=ifelse(is.na(tomatometer), -1, tomatometer)) %>%
  ggplot(aes(x = tomatometer, y = fct_reorder(title, tomatometer))) +
    geom_point(color = "blue") +
    scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    theme_dotplot +
    xlab("Tomatometer score") +
    ylab("") +
    ggtitle("Movies ranked by tomatometer score")
```

(d) Create a Cleveland dot plot of tomatometer *and* audience scores on the same graph, one color for each. Sort by audience score.
```{r fig.height=4, fig.width=3}

movies %>%
  # Replace NA score with -1, which causes those movies to be sorted at the bottom of the plot
  # The value of -1 is not shown since we wet the scale for x to be [0,100]
  mutate(tomatometer=ifelse(is.na(tomatometer), -1, tomatometer)) %>%
  mutate(audience_score=ifelse(is.na(audience_score), -1, audience_score)) %>%
  gather(key = "Score", value = "value", tomatometer, audience_score) %>%
  ggplot(aes(x = value, y = fct_reorder2(title, Score=='audience_score', value, .desc = FALSE), color = Score)) +
    geom_point() +
    scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    theme_dotplot + theme(
      legend.position = "top",
      legend.title = element_text(size = rel(0.75))
    ) + 
    xlab("") +
    ylab("") +
    ggtitle("Movies scores")
```


(e) Run your code again for the weekend of July 5 - July 7, 2019. Use **plotly** to create a scatterplot of audience score vs. tomatometer score with the ability to hover over the point to see the film title.
```{r}
movies_july05_july07 <- rottentomatoes.get_movies(rottentomatoes.box_office_urls(rank_id=12))
```

```{r}
g <- ggplot(movies_july05_july07, aes(audience_score, tomatometer, text = paste("title:", title))) +
      geom_point(color = "blue", alpha = .4, na.rm = TRUE) + 
      expand_limits(x = 0, y = 0) + 
      ggtitle("Movie scores") +
      theme_dotplot
ggplotly(g, height = 500, width = 500)
```

### 3. Weather

[14 points]

Data: `weather` dataset in **nycflights13** package (available on CRAN)
```{r}
library(nycflights13)
```

For parts (a) - (d) draw four plots of `wind_dir` vs. `humid` as indicated. For all, adjust parameters to the levels that provide the best views of the data.

(a) Points with alpha blending

```{r}
weather %>%
  #sample_n(10000) %>%
  ggplot(aes(wind_dir, humid)) +
  geom_point(color = "blue", size = 1.5, alpha = .1, stroke = 0, position = "jitter") + 
  scale_x_continuous(limits = c(0, 360), breaks = seq(0, 360, 30)) +
  #coord_polar(theta = "x", direction = 1, start = 0) +
  xlab("wind direction") +
  ylab("humidity") +
  ggtitle("Scatterplot of wind/humididy of weather")
```

(b) Points with alpha blending + density estimate contour lines

```{r}
weather %>%
  ggplot(aes(wind_dir, humid)) +
  geom_point(color = "blue", size = 1.5, alpha = .1, stroke = 0, position = "jitter") + 
  scale_x_continuous(limits = c(0, 360), breaks = seq(0, 360, 30)) +
  geom_density2d(bins = 10, color = "black") +
  #coord_polar(theta = "x", direction = 1, start = 0) +
  xlab("wind direction") +
  ylab("humidity") +
  ggtitle("Scatterplot + density of wind/humididy of weather")
```

(c) Hexagonal heatmap of bin counts

```{r}
weather %>%
  ggplot(aes(wind_dir, humid)) +
  scale_x_continuous(limits = c(0, 360), breaks = seq(0, 360, 30)) +
  geom_hex(binwidth = c(20, 10)) +
  scale_fill_viridis_c() +
  #coord_polar(theta = "x", direction = 1, start = 0) +
  xlab("wind direction") +
  ylab("humidity") +
  ggtitle("Hexagonal heatmap of wind/humididy of weather")
```

(d) Square heatmap of bin counts 

```{r}
weather %>%
  ggplot(aes(wind_dir, humid)) +
  scale_x_continuous(limits = c(0, 360), breaks = seq(0, 360, 30)) +
  geom_bin2d(binwidth = c(10, 5)) +
  scale_fill_viridis_c() +
  #coord_polar(theta = "x", direction = 1, start = 0) +
  xlab("wind direction") +
  ylab("humidity") +
  ggtitle("Square heatmap of wind/humididy of weather")
```

(e) Describe noteworthy features of the data, using the "Movie ratings" example on page 82 (last page of Section 5.3) as a guide.  

(f) Draw a scatterplot of `humid` vs. `temp`. Why does the plot have diagonal lines?

```{r}
weather %>%
  ggplot(aes(humid, temp)) +
  geom_point(color = "blue", alpha = .1, stroke = 0) + 
  xlab("humidity") +
  ylab("temperatur") +
  ggtitle("Scatterplot of humidity/temperature of weather")
```

(g) Draw a scatterplot matrix of the continuous variables in the `weather` dataset. Which pairs of variables are strongly positively associated and which are strongly negatively associated?

```{r fig.height=7, fig.width=7}
library(lattice)
weather %>% 
  select(month, temp, dewp, humid, precip, pressure, visib, wind_dir, wind_speed) %>% 
  sample_n(1000) %>%
  splom()
```
Strong positive correlation exists between `temp` and `dewp` variables, the higher temperature the higher dew point, and the lower the temperature, the lower the dew point. This is a well known physics phenomenon.

Negative correlation can be noticed between humidity and visibility. As humidity increases, the weather becomes foggy or rainy, which decreases visibility.


(h) Color the points by `origin`.  Do any new patterns emerge?

```{r fig.height=7, fig.width=7}
w <- weather %>% 
  mutate(origin = as.factor(origin)) %>%
  select(origin, month, temp, dewp, humid, precip, pressure, visib, wind_dir, wind_speed) %>% 
  sample_n(500)
my_cols <- c("red", "blue", "green")
splom(w, col = my_cols[w$origin])
```

